\documentclass{article}
\usepackage{listings}

\title{Working Distributions}
\author{Drew Dolgert}
\date{\today}

\newcommand{\code}[1]{\texttt{#1}}

\begin{document}
\maketitle

\section{Introduction}
This describes implementation of distributions for sampling of
distributions for stochastic simulation in continuous time.
This kind of simulation makes specific demands on what calls
a distribution must support, and those calls are different from
what libraries provide.



\section{Shifted Absolute Distributions}
Our mission is to express probability distributions
with respect to some enabling time $t_e$ and then, after
some later time $t_0$, to perform three operations, sampling,
integration of the hazard, and the inverse of integration of
the hazard.

We know we will have some standard functions on hand from the programming
language libraries. These include the cumulative distribution function.
Our first question is how to express the cumulative distribution
function for a distribution that is shifted in time, so that we aren't
counting from time zero.

Given a cumulative distribution,we normally write
\begin{equation}
  F(t)=1-e^{-\int_{0}^t \lambda(s)ds}\label{eqn:simplecdf}
\end{equation}
or
\begin{equation}
  F(t)=\int_0^t f(s)ds,
\end{equation}
but for distributions in absolute time, meaning they are shifted
by $t_e$. The hazard rate is defined in terms of a zero time,
so we subtract the enabling time.
\begin{equation}
  F(t;t_e)=1-e^{-\int_{t_e}^{t} \lambda(s-t_e)ds}
\end{equation}
Another way to write this is to assume the enabling time is included
in the integral limits.
\begin{equation}
  F(t,t_e)=1-e^{-\int_{0}^{t-t_e} \lambda(s)ds}
\end{equation}
or
\begin{equation}
  F(t,t_e)=\int_{0}^{t-t_e} f(s)ds.
\end{equation}
The call for $F(t)$ is \code{cdf(t)}, and $F(t,t_e)$ is \code{cdf($t-t_e$)}.

Now let's sample the distribution after some time $t_0>=t_e$, at which time
we normalize the distribution to its remaining probability. We can
think of this best with survivals, $G(t)=1-F(t)$. In words, the probability of
survival from $t_e$ to time $t$ ($G(t,t_e)$) is the probability
of survival from $t_e$ to $t_0$ ($G(t_0,t_e$)
\emph{and} the probability of survival from $t_0$ to $t$ ($G(t_0,t_e)$).
\begin{equation}
  G(t,t_e)=G(t_0,t_e)G(t,t_0,t_e)
\end{equation}
Written in terms of hazards, this is
\begin{equation}
  e^{-\int_{0}^{t-t_e} \lambda(s)ds}=e^{-\int_{0}^{t_0-t_e} \lambda(s)ds}
    e^{-\int_{t_0-t_e}^{t-t_e} \lambda(s)ds},
\end{equation}
where the hazard is the same zero-based hazard from Eq.~\ref{eqn:simplecdf}.
Therefore, given the initial survival, expressed since the enabling time $t_e$,
the scaled survival is
\begin{equation}
  G(t,t_0,t_e)=G(t,t_e)/G(t_0,t_e).
\end{equation}
In terms of cumulative distribution functions,
\begin{eqnarray}
  F(t,t_0,t_e)&=&1-\frac{1-F(t,t_e)}{1-F(t_0,t_e)} \\
          &=&\frac{F(t,t_e)-F(t_0,t_e)}{1-F(t_0,t_e)}\label{eqn:shiftcum} \\
\end{eqnarray}
In terms of the function calls, that means
\begin{lstlisting}
  cdf(t, t0, te)=(cdf(t-te)-cdf(t0-te))/(1-cdf(t0-te))
\end{lstlisting}

\begin{table}
\begin{tabular}{ll} \hline
\code{cdf(d,t)} & $F_d(t)$ \\
\code{quantile(d,q)} & $F_d^{-1}(q)$ \\
\code{logcdf(d,t)} & $\ln(F_d(t))$ \\
\code{ccdf(d,t)} & $G_d(t)$ \\
\code{logccdf(d,t)} & $-\int_0^t \lambda_d(s)ds$ \\
\code{quantile(d,q)} & $F_d^{-1}(q)$ \\
\code{cquantile(d,q)} & $F_d^{-1}(1-q)=G_d^{-1}(q)$ \\
\code{invlogcdf(d,lp)} & $F_d^{-1}(e^{l_p})$ \\
\code{invlogccdf(d,lp)} & $G_d^{-1}(e^{l_p})$ or $-\int_0^{t(l_p)}\lambda(s)ds=l_p$ \\\hline
\end{tabular}
\caption{Translation of methods into math.\label{fig:methodmath}}
\end{table}

In practice, sampling algorithms are specific to particular distributions.
They are formally equivalent to drawing a uniform random variable between
0 and 1, which we call $U$, and solving $U=F(t')$ for $t'$.
For the case of interest, where the distribution has an enabling time, $t_e$,
and is being observed after survival to a time $t_0$, sampling is formally
a solution $t'$ to $U=F(t', t_0, t_e)$. Looking back at Eq.~\ref{eqn:shiftcum},
we can write this as
\begin{eqnarray}
 U&=&F(t,t_0,t_e) \\
  &=&\frac{F(t,t_e)-F(t_0,t_e)}{1-F(t_0,t_e)} \\
U(1-F(t_0,t_e))&=&F(t,t_e)-F(t_0,t_e) \\
F(t,t_e)&=&U(1-F(t_0,t_e))+F(t_0,t_e) \\
F(t-t_e)&=&U(1-F(t_0-t_e))+F(t_0-t_e) \\
t-t_e &=& F^{-1}\left[U(1-F(t_0-t_e))+F(t_0-t_e)\right] \\
t &=& t_e+F^{-1}\left[U(1-F(t_0-t_e))+F(t_0-t_e)\right]
\end{eqnarray}
Using the inverse \textsc{CDF} from Table~\ref{fig:methodmath},
the inverse of this shifted quantile is
\begin{lstlisting}
  quantile(U, t0, te)=te+quantile(U+(1-U)*cdf(t0-te))
\end{lstlisting}
This would be a way to sample any distribution with a \textsc{CDF} and
quantile, but likely badly.

The next two pieces concern the hazard. The goal is to find the integral
of the hazard between two absolute times, $t_1$ and $t_2$, where both
are $t_{1,2}\ge t_0$. This is
\begin{equation}
  \int_{t_1-t_e}^{t_2-t_e} \lambda(s)ds=\int_{0}^{t_2-t_e} \lambda(s)ds
  	-\int_{0}^{t_1-t_e} \lambda(s)ds.
\end{equation}
In terms of the given methods, this would be, noting the minus sign
in the table,
\begin{lstlisting}
  hazard_int(t1, t2, te)=logccdf(t1-te)-logccdf(t2-te)
\end{lstlisting}

Last is the inverse hazard. We want to solve for $t'$ in
\begin{equation}
  x=\int_{t_0-t_e}^{t'-t_e}\lambda(s)ds.
\end{equation}
Expanding as before, this is
\begin{eqnarray}
  x&=&\int_{0}^{t'-t_e}\lambda(s)ds-\int_{0}^{t_0-t_e}\lambda(s)ds \\
  x+\int_{0}^{t_0-t_e}\lambda(s)ds&=&\int_{0}^{t'-t_e}\lambda(s)ds \\
  -x-\int_{0}^{t_0-t_e}\lambda(s)ds&=& -\int_{0}^{t'-t_e}\lambda(s)ds \\
  l_p&=&-x+\left[-\int_{0}^{t_0-t_e}\lambda(s)ds\right] \\
  l_p&=&-\int_{0}^{t'-t_e}\lambda(s)ds
\end{eqnarray}
Translating this into equations, we get
\begin{lstlisting}
  lp=-x+logccdf(t0-te)
  inv_hazard_int(x, t0, te)=te+invlogccdf(lp)
\end{lstlisting}

\section{Log-Logistic}
Working from wikipedia, because that's smart.
\begin{equation}
  F(x;\alpha, \beta)=\frac{1}{1+(x/\alpha)^{-\beta}}.
\end{equation}
We shift this to
\begin{equation}
  F(t, t_e)=\frac{1}{1+((t-t_e)/\alpha)^{-\beta}}.
\end{equation}
The pdf is
\begin{equation}
  f(x;\alpha, \beta)=\frac{(\beta/\alpha)(x/\alpha)^{\beta-1}}
  {(1+(x/\alpha)^\beta)^2}.
\end{equation}
The quantile is
\begin{equation}
  F^{-1}(p; \alpha, \beta)=\alpha \left(\frac{p}{1-p}\right)^{1/\beta}.
\end{equation}
Survival
\begin{equation}
  G(t)=1-F(t)=\frac{1}{1+(t/\alpha)^\beta}.
\end{equation}
Hazard
\begin{equation}
  \lambda(t)=\frac{f(t)}{G(t)}=\frac{(\beta/\alpha)(t/\alpha)^{\beta-1}}
  {1+(t/\alpha)^\beta}
\end{equation}
Lastly, we need \code{invlogccdf(d,lp)}, which is $G_d^{-1}(e^{l_p})$,
or $-\int_0^t(l_p)\lambda(s)ds=l_p$.
\begin{eqnarray}
  l_p&=&\ln(G(t)) \\
  e^{l_p}&=&G(t) \\
  e^{l_p}&=&\frac{1}{1+(t/\alpha)^\beta} \\
  e^{-l_p}&=&1+(t/\alpha)^\beta \\
  (t/\alpha)^\beta&=&  1-e^{-l_p}\\
  t/\alpha&=& (1-e^{-l_p})^{1/\beta}\\
   t&=&\alpha(1-e^{-l_p})^{1/\beta}\\
\end{eqnarray}

\section{Gamma}
We will define paramaters from the shape $\alpha$ and rate $\beta$.
\begin{equation}
  f(x)=\frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-\beta x}
\end{equation}
where
\begin{equation}
  \Gamma(t)=\int_0^\infty x^{t-1}e^{-x}dx.
\end{equation}
The CDF is
\begin{equation}
  F(x;\alpha,\beta)=\frac{\gamma(\alpha,\beta x)}{\Gamma(\alpha)}
\end{equation}
where $\gamma$ is the (lower) incomplete gamma function,
\begin{equation}
  \gamma(x;\alpha)=\int_0^x t^{\alpha-1}e^{-t}dt
\end{equation}

In our back pocket, from \texttt{Boost::Math}, are
$\Gamma(x)$, $\ln(|\Gamma(x)|)$, digamma, which is
\begin{equation}
  \psi(x)=\frac{d}{dx}\ln(\Gamma(x))=\frac{\Gamma'(x)}{\Gamma(x)},
\end{equation}
gamma ratio, which is $\Gamma(a)/\Gamma(b)$,
gamma delta ratio, which is $\Gamma(a)/\Gamma(a+\Delta)$,
and the set of incomplete gamma functions.
In order, they are normalized lower incomplete, normalized upper, incomplete
full (non-normalized) lower incomplete, and full (non-normalized)
upper incomplete gamma functions.
\begin{eqnarray}
  \mbox{gamma\_p}(a,z)&=&\frac{\gamma(a,z)}{\Gamma(a)}=\frac{1}{\Gamma(a)}
     \int_0^zt^{a-1}e^{-t}dt \\
  \mbox{gamma\_q}(a,z)&=&\frac{\Gamma(a,z)}{\Gamma(a)}=\frac{1}{\Gamma(a)}
     \int_z^0t^{a-1}e^{-t}dt \\
  \mbox{tgamma\_lower}(a,z)&=&\gamma(a,z)=
     \int_0^zt^{a-1}e^{-t}dt \\
  \mbox{tgamma}(a,z)&=&\Gamma(a,z)=\frac{1}{\Gamma(a)}
     \int_z^0t^{a-1}e^{-t}dt \\
\end{eqnarray}
There are a set of inverses of incomplete gamma functions
and derivatives of incomplete gamma functions.
OK, back to what we need.
\begin{eqnarray}
  F(x;\alpha,\beta)&=&\mbox{gamma\_p}(\alpha, \beta x) \\
  F^{-1}(y;\alpha,\beta)&=&\mbox{gamma\_p\_inv}(\alpha, y)/\beta
\end{eqnarray}

The hazard integral, in terms of the cdf, is
\begin{eqnarray}
 \int_{t_1-t_e}^{t_2-t_e}\lambda(s)ds&=&-\ln(1-F(t_2-t_e))+\ln(1-F(t_1-t_e)) \\
 &=& \ln\left[\frac{1-F(t_1-t_e)}{1-F(t_2-t_e)}\right].
\end{eqnarray}
Can we simplify this into something provided?
\begin{eqnarray}
\int_{t_1-t_e}^{t_2-t_e}\lambda(s)ds & = & \ln\left[\frac{1-\frac{\gamma(\alpha,\beta (t_1-t_e))}{\Gamma(\alpha)}}{1-\frac{\gamma(\alpha,\beta (t_2-t_e))}{\Gamma(\alpha)}}\right] \\
 & = & \ln\left[\frac{\Gamma(\alpha)-\gamma(\alpha,\beta (t_1-t_e))}
 {\Gamma(\alpha)-\gamma(\alpha,\beta (t_2-t_e))} \right] \\
\gamma(\alpha,\beta (t_1-t_e)) & = & \int_0^{\beta(t_1-t_e)} t^{\alpha-1}e^{-t}dt
\end{eqnarray}
It looks like we might do best just with
\begin{lstlisting}
Ga=tgamma(a)
hazint(te, t1, t2)=log((Ga-tgamma_lower(a,b*(t1-te)))/
    (Ga-tgamma_lower(a,b*(t2-te))))
\end{lstlisting}

Our other goal for Gamma distributions is to get the inverse hazard.
This can be seen as two steps. First find the integral
\begin{equation}
  l_p=-x+\left[\int_0^{t0-t_e}\lambda(s)ds\right].
\end{equation}
Then solve for $t'$ in
\begin{equation}
  l_p=-\int_0^{t'-t_e}\lambda(s)ds.
\end{equation}
Or, we could write this as
\begin{equation}
  l_e =e^{-x}e^{-\int_0^{t0-t_e}\lambda(s)ds}=e^{-x}(1-F(t_0-t_e))
\end{equation}
and
\begin{equation}
  l_e=e^{-\int_0^{t'-t_e}\lambda(s)ds}=1-F(t'-t_e).
\end{equation}
All at once,
\begin{eqnarray}
  F(t'-t_e)&=&1-e^{-x}(1-F(t_0-t_e)) \\
 t'&=&t_e+F^{-1}\left(1-e^{-x}(1-F(t_0-t_e))\right). \\
 F(t_0-t_e)&=&\mbox{gamma\_p}(\alpha,\beta(t_0-t_e)) \\
 F^{-1}(y)&=&\mbox{gamma\_p\_inv}(\alpha, y)/\beta
\end{eqnarray}
So here is our inverse hazard integral.
\begin{lstlisting}
  quad=1-exp(-x)*(1-gamma_p(a,b*(t0-te)))
  tp=te + gamma_p_inv(a, quad)/b
\end{lstlisting}

\section{Uniform Distribution}
Maybe this one will be easier.
This distribution has two parameters, a start time
and an end time, $t_a$ and $t_b$.
The pdf is constant, $f(t)=1/(t_b-t_a)$ between
$t_a\le t<t_b$. The CDF is just the integral of
that, $F(t)=(t-t_a)/(t_b-t_a)$.
The integrated hazard will have nonzero cases for
for $t_1<t_a<t_2<t_b$, $t_1<t_a<t_b<t_2$,
$t_a<t_1<t_2<t_b$, $t_a<t_1<t_b<t_2$.
It is zero for $t_1<t_2<t_a$ and $t_a<t_b<t_1<t_2$
\begin{equation}
  \int_{t_1-t_e}^{t_2-t_e}\lambda(s)ds=
      \ln\left[\frac{1-F(t_1-t_e)}{1-F(t_2-t_e)}\right]
\end{equation}
If $t_a\le t_n-t_e<t_b$, then $F(t_n-t_e)=(t_n-t_e-t_a)/(t_b-t_a)$.
Otherwise it is $0$ or $1$. It should never be the
case that a uniform distribution does not fire
before $t_b$. The hazard integral always sums over
time already past in the simulation. Nevertheless, it will
be necessary to check for overflow near $t_b$, and it
would help to keep the two logs separated, instead of
in the fraction.

What about the inverse of the hazard integral?
$F^{-1}(x)=t_a+(t_b-t_a)x$ Therefore, for $t_a\le t_0-t_e$,
\begin{equation}
  t'=t_e+t_a+(t_b-t_a)\left[1-e^{-x}\left(1-\frac{t_0-t_e-t_a}{t_b-t_a}\right)\right]
\end{equation}
and for $t_0-t_e< t_a$,
\begin{equation}
  t'=t_e+t_a+(t_b-t_a)\left[1-e^{-x}\right]
\end{equation}

\section{Triangular Distribution}
The cumulative distribution function for the triangular distribution
with endpoints $a$ and $b$ and midpoint $m$ is
\begin{eqnarray}
  \frac{(t-a)^2}{(b-a)(m-a)} & & a\le t \le m \\
  1-\frac{(b-t)^2}{(b-a)(b-m)} & & m<t\le b.
\end{eqnarray}
This makes the survival
\begin{eqnarray}
  1-\frac{(t-a)^2}{(b-a)(m-a)} & & a\le t \le m \\
  \frac{(b-t)^2}{(b-a)(b-m)} & & m<t\le b.
\end{eqnarray}
\subsection{Simple Sample}
The cutoff is at $t=m$, which is
\begin{eqnarray}
  U'&=&\frac{(m-a)^2}{(b-a)(m-a)} \\
  &=&\frac{m-a}{b-a}
\end{eqnarray}
so first check whether $U$ is greater than that. Then, for $U$
less than that,
\begin{eqnarray}
  t = a + \left[U(b-a)(m-a)\right]^{1/2} & & U\le U' \\
  t = b- \left[(1-U)(b-a)(b-m)\right]^{1/2} & & U'<U \\
\end{eqnarray}

\subsection{Shifted Sample}
If this is sampled after some time, $x$, then the thing
we want to invert is
\begin{equation}
  U=\frac{F(t)-F(x)}{G(x)}
\end{equation}
so
\begin{equation}
  F(t)=UG(x)+F(x)
\end{equation}
which, for $a<t\le m$ and $a<x\le m$, is
\begin{eqnarray}
  \frac{(t-a)^2}{(b-a)(m-a)} & = & U\left[1-\frac{(x-a)^2}{(b-a)(m-a)}\right]
  +\frac{(x-a)^2}{(b-a)(m-a)} \\
  (t-a)^2 & = & U(b-a)(m-a)+ (1-U)(x-a)^2 \\
  t & = & a + \left[U(b-a)(m-a)+ (1-U)(x-a)^2\right]^{1/2}
\end{eqnarray}
For $m<t\le b$ and $m<x\le b$, this is
\begin{eqnarray}
  1-\frac{(b-t)^2}{(b-a)(b-m)} & = & U\frac{(b-x)^2}{(b-a)(b-m)}+
    1-\frac{(b-x)^2}{(b-a)(b-m)} \\
  -(b-t)^2 & = & U(b-x)^2-(b-x)^2 \\
  t & = & b-(b-x)\sqrt{1-U}
\end{eqnarray}
In the case that $m<t\le b$ and $a<x\le m$, the result is
\begin{eqnarray}
  1-\frac{(b-t)^2}{(b-a)(b-m)} & = & U\left[1-\frac{(x-a)^2}{(b-a)(m-a)}\right]
   + \frac{(x-a)^2}{(b-a)(m-a)}\\
  1-\frac{(b-t)^2}{(b-a)(b-m)} & = &U+\frac{(x-a)^2(1-U)}{(b-a)(m-a)} \\
  \frac{(b-t)^2}{(b-a)(b-m)} & = &(1-U)-\frac{(x-a)^2(1-U)}{(b-a)(m-a)} \\
  (b-t)^2& = &(1-U)(b-a)(b-m)-\frac{(x-a)^2(1-U)(b-m)}{(m-a)} \\
  t & = & b-\left[(1-U)(b-a)(b-m)-\frac{(x-a)^2(1-U)(b-m)}{(m-a)}\right]^{1/2}
\end{eqnarray}

\subsection{Sampling from Quantile}
The equation we have to solve is
\begin{equation}
  F(t)=1-G(t_j)(1-u)\prod_i\frac{G_i(t_i)}{G_i(t_{i+1})},
\end{equation}
given to us as
\begin{equation}
  F(t)=1-G(t_j)(1-u)\gamma,
\end{equation}
so, in terms of survivals, it's
\begin{equation}
  G(t)=G(t_j)(1-u)\gamma
\end{equation}
The value on the right is all known.
If $t_j<m$, the cutoff is
\begin{equation}
  G'=\frac{b-m}{(b-a)(b-m)}
\end{equation}
Below that,

\end{document}
